T-reX: A tool to quantify waste and material footprints in current and future Life Cycle Assessment (LCA) databases

### 2024-03-08 latex_files/introduction.tex
small update to V2

To facilitate the quantification of waste and material flows in LCA, we have developed a Python package built to extend the \texttt{brightway framework}~\citep{mutel2017brightway} and designed to track these exchanges by translating them into inventory indicators and `pseudo' LCA impact (LCIA) categories. T-reX enables LCA practitioners to easily manipulate their databases to allow them to easily aggregate the mass and volume of any desired exchange, and to create flexible categories that differentiate between material categories, waste types, and EOL handling. While some methods with similar aims exist, they lack customisability and specificity~\citep{foen2021ecofactors} or can be cumbersome to apply and suffer from errors due to multiple counting~\citep{laurenti2023wastefootprint}.

>>>>

To facilitate the quantification of waste and material flows in LCA, we developed a Python package to extend the \texttt{brightway framework}~\citep{mutel2017brightway} and designed to track these exchanges by translating them into inventory indicators and `pseudo' LCA impact (LCIA) categories. T-reX means for `Tool for re-X' (reduce, reuse, recycle, etc.).  T-reX enables LCA practitioners to easily manipulate their databases to allow them to easily aggregate the mass and volume of any desired exchange, and to create flexible categories that differentiate between material categories, waste types, and EOL handling. While some methods with similar aims exist, they lack customisability and specificity~\citep{foen2021ecofactors} or can be cumbersome to apply and suffer from errors due to multiple counting~\citep{laurenti2023wastefootprint}.

### 2024-03-08 latex_files/main.tex
small update to V2

\linenumbers
% add something about the future stuffg
    % SPOTLIGHTS (5 bullet points, max 120 characters each)
    {\Large Spotlights}
    \vspace{1em}
    \begin{description}[style=nextline]
        \item[Bullet 1: Critical context and background information on the problem addressed] Tracking waste and material flows is essential for developing the circular economy and supply chain resilience
        \item[Bullet 2: A brief overview of the key finding of the study (or findings if necessary)] A tool was developed to quantify waste and material footprints of activities in life cycle assessment (LCA) databases
        \item[Bullet 3: The most radical, creative, disruptive or innovative aspect of the manuscript] Built to complement the Brightway LCA framework and ActivityBrowser, the tool is customisable and easy to use.
        \item[Bullet 4: The significance of the results to the environment, economics, or society] The tool can be used to identify pre-consumer waste and material hotspots that are often hidden in supply chains.
        \item[Bullet 5: Future vision or the most important implications for continued research] Improved data availability and quality would enable more detailed and accurate waste and material footprinting.
    \end{description}

@@

    %% RESEARCH HIGHLIGHTS (3-5 bullet points, max 85 characters each)
    \begin{highlights}
        \item T-reX, a new tool for quantifying waste and material flows in LCA.
        \item Assesses supply risks by calculating demand for critical materials.
        \item Simplifies quantification of user-specified waste and material categories.
        \item Rapidly identifies waste and material demand hotspots.
        \item Presents a case study of five Li-ion battery supply chains`'.
    \end{highlights}

@@

        A case study of 5 Li-ion batteries demonstrated T-reX's utility, quantifying categorised waste and material inventory footprints, and thus, their potential environmental burdens. T-reX can aid sustainable decision-making and contribute to the development of the `circular economy' by facilitating analysis of material consumption and waste generation in LCA.

@@

\end{frontmatter}
\cbox{Total word count: ca. 5200, Limit: 5000}

% list of abbreviations
\begin{table}[H]extension
    \centering
    \caption{List of abbreviations}
    \label{tab:abbreviations}
    \begin{tabular}{ll}
    \toprule
    CF & Carbon Footprint \\
    EF & Ecological Footprint \\
    CML & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
    CRM & Critical Raw Material \\
    CPC & Cooperative Patent Classification \\
    CSI & Crustal Scarcity Indicator \\
    CSP & Crustal Scarcity Potential \\
    EDIP & Environmental Design of Industrial Products \\
    EOL & End-of-Life \\
    FOEN & Swiss Federal Office for the Environment \\
    IAM & Integrated Assessment Model \\
    LCA & Life Cycle Assessment \\
    LCI & Life Cycle Inventory \\
    LCIA & Life Cycle Impact Assessment \\
    LFP & Lithium Iron Phosphate \\
    LiMn$_2$O$_4$ & Lithium Manganese Oxide \\
    MF & Material Footprint \\
    NCA & Nickel Cobalt Aluminum \\
    NMC & Nickel Manganese Cobalt \\
    PWF & Product Waste Footprint \\
    RCP & Representative Concentration Pathway \\
    ReCiPe & A standard LCIA method set \\
    SSPs & Shared Socioeconomic Pathways \\
    UBP & Umweltbelastungspünkte (Environmental Impact Points) \\
    WF & Waste Footprint \\
    \bottomrule
    \end{tabular}
    \end{table}

%% main text
\section{Introduction}
\label{sec:introduction}
\cbox{Section word count: 1700}
\cbox{To be split into two (sub) sections: Introduction and Background, can be easily condensed by 200 words or so.}
\input{introduction.tex}

@@

\section{Discussion}
\label{sec:discussion}
\input{discussion.tex}
\cbox{Section word count: 450}

\section{Conclusions}
\label{sec:conclusions}
\cbox{Section word count: 150}
We have created the T-reX tool, an extension to the brightway LCA ecosystem that enables users to calculate the waste and material footprints of a product or service in an LCA database. It explodes the database, identifies upstream waste and material exchanges, edits them, and writes matching custom `pseudo' LCIA methods. These exchanges become pseudo-biosphere flows and thus, the footprint can be calculated as with the existing LCIA methods. The T-reX tool can be easily customised by the user to calculate the footprints of other supply chain flows such as water, gas, and critical raw materials.

@@

% Back matter
\section*{Data availability} 
All data used in this analysis are publicly available online under the noted sources. The T-reX tool is installable via the Python Package Index (PyPI) and is available at \url{https://pypi.org/project/Tool-reX}.
The full source code for the T-reX tool is available at \url{https://www.github.com/Stew-McD/T-reX}.
A user guide and comprehensive documentation are available at \url{https://Tool-reX.readthedocs.io}.

@@

>>>>

\linenumbers
% add something about the future stuffg
    % SPOTLIGHTS (5 bullet points, max 120 characters each)
    {\Large Spotlights}
    \vspace{1em}
    \begin{description}[style=nextline]
        \item[Bullet 1: Critical context and background information on the problem addressed] Tracking waste and material flows is essential for developing the circular economy and supply chain resilience
        \item[Bullet 2: A brief overview of the key finding of the study (or findings if necessary)] A tool was developed to quantify waste and material footprints of activities in life cycle assessment (LCA) databases
        \item[Bullet 3: The most radical, creative, disruptive or innovative aspect of the manuscript] Built to complement the Brightway LCA framework and ActivityBrowser, the tool is customisable and easy to use
        \item[Bullet 4: The significance of the results to the environment, economics, or society] The tool can be used to identify pre-consumer waste and material hotspots that are often hidden in supply chains
        \item[Bullet 5: Future vision or the most important implications for continued research] Improved data availability and quality would enable more detailed and accurate waste and material footprinting
    \end{description}

@@

    %% RESEARCH HIGHLIGHTS (3-5 bullet points, max 85 characters each)
    \begin{highlights}
        \item T-reX, a new tool for quantifying waste and material flows in LCA
        \item Assesses supply risks by calculating demand for critical materials
        \item Simplifies quantification of user-specified waste and material categories
        \item Rapidly identifies waste and material demand hotspots
        \item Presents a case study of five Li-ion battery supply chains
    \end{highlights}

@@

        A case study of five Li-ion batteries demonstrated T-reX's utility, quantifying categorised waste and material inventory footprints, and thus, their potential environmental burdens. T-reX can aid sustainable decision-making and contribute to the development of the `circular economy' by facilitating analysis of material consumption and waste generation in LCA.

@@

\end{frontmatter}
\cbox{Total word count: ca. 5200, Limit: 5000}
\cbox{Note for reviewers: please annotate the pdf and email it to me at \href{mailto:sc.mcdowall@cml.leidenuniv.nl}{sc.mcdowall@cml.leidenuniv.nl}}
\cbox{Supplementary material is available \href{https://github.com/Stew-McD/T-reX_Publication/blob/main/reviews/T-reX_Manuscript_V2/T-reX_manuscript_V2_SupplementaryMaterial.pdf}{here}}
\cbox{Style guide for the journal: \href{https://www.sciencedirect.com/journal/resources-conservation-and-recycling/publish/guide-for-authors}{here}}

% list of abbreviations
\begin{table}[H]
    \centering
    \caption{List of abbreviations}
    \label{tab:abbreviations}
    \begin{tabular}{ll}
    \toprule
    CF & Carbon Footprint \\
    EF & Ecological Footprint \\
    CML & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
    CRM & Critical Raw Material \\
    CPC & Cooperative Patent Classification \\
    CSI & Crustal Scarcity Indicator \\
    CSP & Crustal Scarcity Potential \\
    EDIP & Environmental Design of Industrial Products \\
    EOL & End-of-Life \\
    FOEN & Swiss Federal Office for the Environment \\
    IAM & Integrated Assessment Model \\
    LCA & Life Cycle Assessment \\
    LCI & Life Cycle Inventory \\
    LCIA & Life Cycle Impact Assessment \\
    LFP & Lithium Iron Phosphate \\
    LiMn$_2$O$_4$ & Lithium Manganese Oxide \\
    MF & Material Footprint \\
    NCA & Nickel Cobalt Aluminum \\
    NMC & Nickel Manganese Cobalt \\
    PWF & Product Waste Footprint \\
    RCP & Representative Concentration Pathway \\
    ReCiPe & A standard LCIA method set \\
    SSPs & Shared Socioeconomic Pathways \\
    UBP & Umweltbelastungspünkte (Environmental Impact Points) \\
    WF & Waste Footprint \\
    \bottomrule
    \end{tabular}
    \end{table}

%% main text
\section{Introduction}
\label{sec:introduction}
\cbox{Section word count: 1700}
\input{introduction.tex}

@@

\section{Discussion}
\label{sec:discussion}
\cbox{Section word count: 450}
\input{discussion.tex}

\section{Conclusions}
\label{sec:conclusions}
\cbox{Section word count: 150}
We have created the T-reX tool, an extension to the brightway LCA ecosystem that enables users to calculate the waste and material footprints of a product or service in an LCA database. It explodes the database, identifies upstream waste and material exchanges, edits them, and writes matching custom `pseudo' LCIA methods. These exchanges become pseudo-biosphere flows and thus, the footprint can be calculated as with the existing LCIA methods. T-reX can be easily customised by the user to calculate the footprints of other supply chain flows such as water, gas, and critical raw materials.

@@

% Back matter
\section*{Data availability} 
All data used in this analysis are publicly available online under the noted sources.T-reX is installable via the Python Package Index (PyPI) and is available at \url{https://pypi.org/project/T_reX_LCA}.
The full source code for the T-reX tool is available at \url{https://www.github.com/Stew-McD/T-reX}.
A user guide and comprehensive documentation are available at \url{https://T-reX.readthedocs.io}.

@@

\begin{itemize}[noitemsep]
    \item \textbf{SCM}: Stewart Charles McDowall
    \item \textbf{EL}: Elizabeth Lanphear
    \item \textbf{SC}: Stefano Cucurachi
    \item \textbf{CFB}: Carlos Felipe Blanco
\end{itemize}

### 2024-03-08 latex_files/methodology.tex
small update to V2

This workflow creates copy of the original \texttt{brightway} project containing the original biosphere database, a T-reX 'pseudo-biosphere' database along with one or more manipulated technosphere database(s) that can be used to calculate the waste and material inventory footprints of activities in the same way as standard LCIA calculations.

@@

\begin{figure}[H]
    \centering
    \caption{Workflow of the T-reX. Application of T-reX to one or more LCA databases creates a brightway project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
    \includegraphics[width=0.5\linewidth]{figures/T-reX_method.pdf}
    \label{fig:methods-flowchart}
\end{figure}

@@

The source of the life cycle inventory data for this case study was ecoinvent 3.9.1 cutoff. Additionally, the T-reX tool was used to create prospective database sets using the \mbox{REMIND} model with the following Representative Concentration Pathways (RCPs):
\begin{itemize}
    \item SSP2-base: representing an approximate 3.5°C increase in global temperatures to 2100
    \item SSP2-PkBudg500: representing the achievement of Paris climate goals, ca. 1.3°C increase to 2100
\end{itemize}

>>>>

This workflow creates copy of the original \texttt{brightway} project containing the original biosphere database, a T-reX `pseudo-biosphere' database along with one or more manipulated technosphere database(s) that can be used to calculate the waste and material inventory footprints of activities in the same way as standard LCIA calculations.

@@

\begin{figure}[H]
    \centering
    \caption{Workflow of the T-reX. Application of T-reX to one or more LCA databases creates a brightway project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain. [TO BE IMPROVED!]}
    \includegraphics[width=0.5\linewidth]{figures/T-reX_method.pdf}
    \label{fig:methods-flowchart}
\end{figure}

@@

The source of the life cycle inventory data for this case study was the database ecoinvent, version 3.9.1 with the `cutoff' attributional system model. Additionally, T-reX was used to create prospective database sets using the \mbox{REMIND} model and the baseline scenario SSP2 `middle of the road' with the following Representative Concentration Pathways (RCPs):
\begin{itemize}
    \item SSP2-base: representing an approximate 3.5°C increase in global temperatures to 2100
    \item SSP2-PkBudg500: representing the achievement of Paris climate goals, ca. 1.3°C increase to 2100
\end{itemize}

### 2024-03-08 latex_files/results.tex
small update to V2

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/T-reX\_NMC811\_WasteTotalSolid.pdf}
    \caption{A Sankey diagram of the total solid waste footprint flows for the NMC 811 battery at the global market in the ecoinvent database version 3.9.1}\label{fig:sankey_waste}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{figures/T-reX\_NMC811\_WasteTotalSolid.pdf}
    \caption{Sankey diagram depicting the supply chain flows of elemental chromium for 1~kg of the LFP (lithium iron phosphate) battery in the case study at the global market in the ecoinvent database version 3.9.1}\label{fig:sankey_material}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{figures/carbondioxide.png}
    \caption{Carbon dioxide waste (from carbon capture and storage) footprints for the five Li-ion batteries in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:carbondioxide}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{figures/phosphate.png}
    \caption{Phosphate material demand footprints for the five Li-ion batteries in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:phosphate}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/top-processes.png}
    \caption{Contribution of `top-processes' to the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:top_contribution}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/cpc_contribution.png
    }
    \caption{Contribution of industrial sectors to the liquid waste footprint of the NCA battery from 2020 to 2100 under the SSP2 scenario using the PkBudg500 RCP of the REMIND model.}\label{fig:cpc_contribution}
\end{figure}

@@

    \begin{subfigure}[b]{0.8\linewidth}
        \includegraphics[width=\linewidth]{figures/comparison_methods1.png}
        \caption{}\label{fig:comparison_methods_b}
    \end{subfigure}
    \caption{Comparison of LCIA method `EDIP 2003 - coal no LT' (a) with the `T-reX - Coal (black)' (b) in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:comparison_methods}
\end{figure}

>>>>

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/T-reX\_NMC811\_WasteTotalSolid.pdf}
    \caption{A Sankey diagram of the total solid waste footprint flows for the NMC 811 battery at the global market in the ecoinvent database version 3.9.1 [TO BE IMPROVED!]}\label{fig:sankey_waste}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{figures/T-reX\_NMC811\_WasteTotalSolid.pdf}
    \caption{Sankey diagram depicting the supply chain flows of elemental chromium for 1~kg of the LFP (lithium iron phosphate) battery in the case study at the global market in the ecoinvent database version 3.9.1 [TO BE IMPROVED!]}\label{fig:sankey_material}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{figures/carbondioxide.png}
    \caption{Carbon dioxide waste (from carbon capture and storage) footprints for the five Li-ion batteries in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model. [TO BE IMPROVED!]}\label{fig:carbondioxide}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{figures/phosphate.png}
    \caption{Phosphate material demand footprints for the five Li-ion batteries in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model. [TO BE IMPROVED!]}\label{fig:phosphate}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/top-processes.png}
    \caption{Contribution of `top-processes' to the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model. [TO BE IMPROVED!]}\label{fig:top_contribution}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/cpc_contribution.png
    }
    \caption{Contribution of industrial sectors to the liquid waste footprint of the NCA battery from 2020 to 2100 under the SSP2 scenario using the PkBudg500 RCP of the REMIND model. [TO BE IMPROVED!]}\label{fig:cpc_contribution}
\end{figure}

@@

    \begin{subfigure}[b]{0.8\linewidth}
        \includegraphics[width=\linewidth]{figures/comparison_methods1.png}
        \caption{}\label{fig:comparison_methods_b}
    \end{subfigure}
    \caption{Comparison of LCIA method `EDIP 2003 - coal no LT' (a) with the `T-reX - Coal (black)' (b) in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.[TO BE IMPROVED!]}\label{fig:comparison_methods}
\end{figure}
